---
title: "Blood Pressure and Health Characteristics"
author: "Sara Giuffre-Coppersmith"
format:
  dashboard:
    theme: cosmo
---

```{r}
#| include: false

library(tidyverse)
library(knitr)
library(plotly)   # interactive plots
library(DT)       # interactive tables

data <- read_csv("data/nhanes_dashboard.csv", show_col_types = FALSE)

num_cols <- c("RIDAGEYR", "BMXBMI", "BPXOSY1", "BPXOSY2", "BPXOSY3", "systolic_bp")
data <- data %>%
  mutate(across(any_of(num_cols), ~ suppressWarnings(as.numeric(.x))))

n            <- nrow(data)
average_age  <- mean(data$RIDAGEYR, na.rm = TRUE)
average_bmi  <- mean(data$BMXBMI, na.rm = TRUE)
average_bp   <- mean(data$systolic_bp, na.rm = TRUE)

bmi_bp_data <- data %>%
  select(BMXBMI, systolic_bp) %>%
  drop_na()

bmi_bp_cor   <- cor(bmi_bp_data$BMXBMI, bmi_bp_data$systolic_bp)
bmi_bp_model <- lm(systolic_bp ~ BMXBMI, data = bmi_bp_data)
bmi_bp_p     <- summary(bmi_bp_model)$coefficients["BMXBMI", "Pr(>|t|)"]

model_data <- data %>%
  select(systolic_bp, BMXBMI, RIDAGEYR, sex, smoking) %>%
  drop_na()

adj_model  <- lm(systolic_bp ~ BMXBMI + RIDAGEYR + sex + smoking, data = model_data)
adj_coefs  <- summary(adj_model)$coefficients
adj_bmi_est <- adj_coefs["BMXBMI", "Estimate"]
adj_bmi_p   <- adj_coefs["BMXBMI", "Pr(>|t|)"]


smoking_p <- anova(adj_model)["smoking", "Pr(>F)"]

missing_values <- data %>%
  summarise(
    Age         = sum(is.na(RIDAGEYR)),
    BMI         = sum(is.na(BMXBMI)),
    Systolic_BP = sum(is.na(systolic_bp)),
    Sex         = sum(is.na(sex)),
    Race        = sum(is.na(race)),
    Smoking     = sum(is.na(smoking))
  )
```

# Overview

## Row {height=22%}

```{r}
#| content: valuebox
#| title: "Participants"
list(icon = "people", color = "primary", value = format(n, big.mark = ","))
```

```{r}
#| content: valuebox
#| title: "Average age (years)"
list(icon = "calendar", color = "info", value = round(average_age, 1))
```

```{r}
#| content: valuebox
#| title: "Average BMI"
list(icon = "speedometer2", color = "warning", value = round(average_bmi, 1))
```

```{r}
#| content: valuebox
#| title: "Average systolic BP (mmHg)"
list(icon = "heart-pulse", color = "danger", value = round(average_bp, 1))
```

## Row

### Purpose

Using NHANES health data, this dashboard explores patterns between blood pressure and  participant characteristics including age, BMI, sex, race/ethnicity, smoking status, and systolic blood pressure. The goal of this dashboard is to make the  NHANES data easier to understand for a general public-health audience through visualizations, summaries, and hypothesis tests.

### Hypotheses

**Hypothesis 1:** A higher BMI is associated with a higher systolic blood pressure, even after considering age, sex, and smoking status.

**Hypothesis 2:** Systolic blood pressure changes due to smoking status after considering age, sex, and BMI.

# Blood Pressure

## Row

### First vs. second reading

```{r}
#| fig-alt: "Comparing first and second systolic blood pressure readings"
p <- ggplot(data, aes(x = BPXOSY1, y = BPXOSY2)) +
  geom_point(alpha = 0.45, color = "#2c7fb8", na.rm = TRUE) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    x = "First systolic BP reading",
    y = "Second systolic BP reading"
  ) +
  theme_minimal()

ggplotly(p)
```

### Distribution of systolic BP

```{r}
#| fig-alt: "Systolic blood pressure across all participants."
ggplot(data, aes(x = systolic_bp)) +
  geom_histogram(bins = 30, fill = "#2c7fb8", color = "white", na.rm = TRUE) +
  labs(x = "Systolic blood pressure (mmHg)", y = "Number of participants") +
  theme_minimal()
```

# Demographics

## Row

### Systolic BP by sex

```{r}
#| fig-alt: "Systolic blood pressure for each sex category."
ggplot(data %>% drop_na(sex, systolic_bp), aes(x = sex, y = systolic_bp, fill = sex)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_viridis_d(option = "D", begin = 0.3, end = 0.8) +
  labs(x = "Sex", y = "Systolic blood pressure (mmHg)") +
  theme_minimal()
```

### Systolic BP by race/ethnicity

```{r}
#| fig-alt: "Systolic blood pressure for each race and ethnicity category."
ggplot(data %>% drop_na(race, systolic_bp), aes(x = race, y = systolic_bp, fill = race)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_viridis_d(option = "D") +
  coord_flip() +
  labs(x = "Race/Ethnicity", y = "Systolic blood pressure (mmHg)") +
  theme_minimal()
```

# BMI and Blood Pressure

## Row {height=22%}

```{r}
#| content: valuebox
#| title: "BMI-BP correlation"
list(icon = "graph-up", color = "primary", value = round(bmi_bp_cor, 2))
```

```{r}
#| content: valuebox
#| title: "Adjusted BMI effect (mmHg per BMI unit)"
list(icon = "activity", color = "info", value = round(adj_bmi_est, 2))
```

## Row

### BMI vs. systolic BP (Hypothesis 1)

```{r}
p <- ggplot(bmi_bp_data, aes(x = BMXBMI, y = systolic_bp)) +
  geom_point(alpha = 0.4, color = "#2c7fb8") +
  geom_smooth(method = "lm", se = TRUE, color = "#d95f0e") +
  labs(x = "BMI", y = "Systolic blood pressure (mmHg)") +
  theme_minimal()

ggplotly(p)
```

### Results for Hypothesis 1

```{r}
h1_results <- data.frame(
  Model = c("Unadjusted", "Adjusted for age, sex, smoking"),
  `BMI coefficient` = round(c(coef(bmi_bp_model)["BMXBMI"], adj_bmi_est), 3),
  `p-value` = c(
    format.pval(bmi_bp_p, digits = 3, eps = 0.001),
    format.pval(adj_bmi_p, digits = 3, eps = 0.001)
  ),
  `Sample size` = c(nrow(bmi_bp_data), nrow(model_data)),
  check.names = FALSE
)

kable(h1_results)
```

## Row

### BMI distribution

```{r}
#| fig-alt: "Body mass index across all participants."
ggplot(data, aes(x = BMXBMI)) +
  geom_histogram(bins = 30, fill = "#2c7fb8", color = "white", na.rm = TRUE) +
  labs(x = "BMI", y = "Number of participants") +
  theme_minimal()
```

# Smoking

## Row {height=22%}

```{r}
#| content: valuebox
#| title: "Hypotheis 2: smoking p-value (adjusted)"
list(
  icon = "clipboard-data",
  color = "primary",
  value = format.pval(smoking_p, digits = 3, eps = 0.001)
)
```

## Row

### Smoking status

```{r}
#| fig-alt: "Number of participants in each smoking status category."
smoking_counts <- data %>%
  filter(!is.na(smoking)) %>%
  count(smoking)

ggplot(smoking_counts, aes(x = smoking, y = n, fill = smoking)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_d(option = "D", begin = 0.2, end = 0.85) +
  labs(x = "Smoking status", y = "Number of participants") +
  theme_minimal()
```

### Smoking status by sex

```{r}
#| fig-alt: "Smoking status counts within each sex category."
smoking_sex_counts <- data %>%
  filter(!is.na(smoking), !is.na(sex)) %>%
  count(sex, smoking)

ggplot(smoking_sex_counts, aes(x = sex, y = n, fill = smoking)) +
  geom_col(position = "dodge") +
  scale_fill_viridis_d(option = "D", begin = 0.2, end = 0.85) +
  labs(x = "Sex", y = "Number of participants", fill = "Smoking status") +
  theme_minimal()
```

## Row

### Systolic BP by smoking status (Hypothesis 2)

```{r}
#| fig-alt: "Systolic blood pressure for each smoking status category."
ggplot(data %>% drop_na(smoking, systolic_bp), aes(x = smoking, y = systolic_bp, fill = smoking)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_viridis_d(option = "D", begin = 0.2, end = 0.85) +
  labs(x = "Smoking status", y = "Systolic blood pressure (mmHg)") +
  theme_minimal()
```

# Explore the Data

## Row

### Participant characteristics

```{r}
age_summary <- data.frame(
  Measure = c(
    "Mean age", "Median age",
    "Mean BMI", "Median BMI",
    "Mean systolic BP", "Median systolic BP"
  ),
  Value = round(
    c(
      mean(data$RIDAGEYR, na.rm = TRUE),   median(data$RIDAGEYR, na.rm = TRUE),
      mean(data$BMXBMI, na.rm = TRUE),     median(data$BMXBMI, na.rm = TRUE),
      mean(data$systolic_bp, na.rm = TRUE), median(data$systolic_bp, na.rm = TRUE)
    ),
    2
  )
)

kable(age_summary)
```

### Data completeness

```{r}
missing_summary <- data.frame(
  Variable = names(missing_values),
  Missing  = as.numeric(missing_values[1, ])
)

kable(missing_summary, col.names = c("Variable", "Missing Values"))
```

## Row

### Browse and filter the participant data

```{r}
# Interactive table: sort, search, and filter each column
data %>%
  select(SEQN, RIDAGEYR, sex, race, smoking, BMXBMI, systolic_bp) %>%
  datatable(
    filter = "top",
    rownames = FALSE,
    options = list(pageLength = 10, scrollX = TRUE)
  )
```

